### 2020-06-17 manuscript.tex
edits 17.6.20

Cryptic species are present throughout the tree of life. They are especially prevalent in ferns, because of processes such hybridization, polyploidy, and reticulate evolution. In addition, the simple morphology of ferns limits phenotypic variation and makes it difficult to detect cryptic species. The model fern genus \textit{Ceratopteris} has long been suspected to harbor cryptic diversity, specifically in the highly polymorphic \textit{C. thalictroides}. Yet no studies have included samples from throughout its pan-tropical range or utilized genomic sequencing, making it difficult to assess the full extent of cryptic variation within this genus. Here, we present the first multilocus phylogeny of the genus using reduced representation genomic sequencing (RADseq) and examine population structure, phylogenetic relationships, and ploidy level variation. We recover similar species relationships found in previous studies, find support for a named cryptic species as genetically distinct, and identify a novel genomic variation within two of the mostly broadly distributed species in the genus.

@@

Despite the recent shift in molecular approaches to investigate cryptic species, much is still unknown about their evolutionary history and ecosystem functions. Cryptic species complexes often have very broad distributions \autocite{Der2009, Knowlton1993, Nygren2014} and can occur in sympatry, allopatry, or parapatry \autocite{Bickford2007}, making adequate sampling challenging. Studies have shown that cryptic species are prevalent across the Tree of Life (e.g. \cite{Bickford2007, Del_Carmen_Molina2011, Hebert2004, Nygren2014}), and ferns in particular are known to harbor many lineages with cryptic variation \autocite{Adjie2007, Paris1989, Yatabe2009}. However, deciphering species boundaries in ferns can be quite difficult due to their ease of dispersal via spores, making gene flow possible across vast distances \autocite{Barrington1993, Tryon1970}. Compared to seed plants, ferns have a high prevalence of polyploidy and reticulate evolution \autocite{Barrington1989, Paris1989, Sigel2016, Otto2000, Otto2000}. In addition, changes in ploidy can contribute to cryptic variation in ferns by altering niche space or offspring viability \autocite{Otto2007, Southgate2019, Masuyama2002}, yet have small effects on phenotype \autocite{Patel2019}. The widespread fern genus \textit{Ceratopteris} Brong. (Pteridaceae) is understood to have considerable cryptic variation, hybridization and polyploidy \autocite{Adjie2007, LloydTax1974, Masuyama2010}, making it an ideal system to investigate the origins and ecology of cryptic species.

@@

Our initial sampling contained 89 specimens of \textit{Ceratopteris} and one sample of \textit{Acrostichum aureum}, plus six replicates. After checking for quality and similarity of replicates, they were combined with the corresponding samples. Of the 90 unique starting samples of \textit{Ceratopteris}, 26 either extracted very poorly or yielded few loci (less than 1000 loci per individual). These were removed from downstream analysis. We also removed 13 samples from Australia, an area we were able to sample on a very fine spatial scale: one individual per Australian population was included in the final dataset so we did not over represent Australian specimens in our analyses. Our final data set included 50 samples of \textit{Ceratopteris} and one sample of \textit{A. aureum}.  We did two separate runs of the \textit{ipyrad} pipeline: one with \textit{A. aureum} for phylogenetic analysis, and one without for ploidy estimation and population structure analysis. Because \textit{A. aureum} is sister to \textit{Ceratopteris} and separated by over 30 million years \autocite{PPGI}, the total number of SNPs retrieved for this individual was quite low.

@@

Due to the presence of numerous cryptic species and morphological variation across the genus, we hypothesize there may be cryptic cytotype variation and/or ploidy type (allo vs. auto) within species, especially the polymorphic \textit{C. thalictroides}. Since a majority our specimens were from herbarium accessions, we could not perform flow cytometry or chromosome squashes. We utilized the R v. 3.5.2 \autocite{R_352} package gbs2ploidy to infer ploidy variation across samples. This package was designed to detect ploidy levels in variable cytoptype populations (specifically quaking aspen, \textit{Populus tremuloides}) using low-coverage (2X) genotyping-by-sequencing (GBS) or RADseq data; it was also tested on simulated data, and can detect diploids, triploids, or tetraploids \autocite{Gompert2017}. This package estimates ploidy from allele ratios, which are calculated from genome-average heterozygosity and bi-allelic SNPs isolated from a variant calling format (VCF) file, which is included in the output from \textit{ipyrad}.

@@

Unfortunately, all samples of \textit{Ceratopteris richardii} (including \textit{Hnn}) either failed in sequencing, or had very poor sequencing quality. After an initial examination of the results of these samples, we chose to remove them from our final analyses. We did not want to attempt to make any inferences about this important species, nor the lab strain, with poor-quality data.

@@

To determin the best \textit{K} value across {\small{STRUCTURE}} we combined several metrics. We used the best \textit{K} method by \textcite{Evanno2005}, which determined a \textit{K} value of 4. The best \textit{K} estimate described in the {\small{STRUCTURE}} manual \autocite{Pritchard2000} determined a \textit{K} value of 5. We also visually examined \textit{K} = 2 - 7 and determined that \textit{K} = 5 was the most biologically meaningful because it separated each named species, and also showed some intra-specific variability (Fig. \ref{structure}). Increasing \textit{K} past 5 did not add any meaningful population clusters.

@@

 The phylogeny resulting from our species quartet inference shows a similar pattern to the {\small{STRUCTURE}} output (Fig. \ref{phy}). Individuals of \textit{Ceratopteris pteridoides} form a monophyletic group, sister to the rest of the genus.z \textit{Ceratopteris cornuta} came out in two places in the phylogeny: sister to Old World \textit{C. thalictroides}, and in a grade outside the \textit{C. gaudichaudii}, Old World and New World \textit{C. thalictroides} clade. Our phylogeny grouped all individuals placed in the \textit{C. gaudichaudii} population by {\small{STRUCTURE}} in a monophyletic clade, sister to New World \textit{C. thalictroides}.

 The {\small{RAxML}} tree shows a very similar structure to the {\small{TETRAD}} tree. The only major difference was that instead of New World \textit{C. thalictroides} and \textit{C. gaudichaudii} forming a monophyletic group, the former comes out as sister to all other species except \textit{C. pteridoides}. Graphical comparison of the two trees is included in Supplementary data 2.

@@

Three of four samples identified as \textit{Ceratopteris cornuta} showed relatively equal allelic proportions of 1:1 and 3:1, indicating a possible tetraploid; the fourth sample was ambiguous. Some samples of the known diploid \textit{C. pteridoides} showed a similar pattern, with one putative diploid and three tetraploids. However, both species are known as a diploid so this may be a mis-assignment of ploidy by our analysis, or detecting of some interesting genomic structure. \textit{Ceratopteris gaudichaudii} had a high 1:1 allele ratio, suggesting that it is an allotetraploid. \textit{Ceratopteris thalictroides} is also known to be a tetraploid \autocite{LloydTax1974, Masuyama2010}, and individuals in our analysis were estimated to have 1:1, 2:1, and 3:1 allele ratios. Despite this variation, there was a distinct pattern between Old World and New World individuals: Old World \textit{C. thalictroides} was most commonly had a 1:1 allele ratio (autopolyploid), whereas New World \textit{C. thalictroides} had  3:1 allele ratios (allopolyploid).

@@

We attempted to create a pseudo-individual to include in our analysis using the \textit{Ceratopteris richardii} genome assembly. We were able to obtain 9,047,554 simulated raw sequences utilizing an \textit{in silico} RADseq digest. These reads, in combination with the rest of our \textit{in vivo} data, were processed with the \textit{ipyrad} pipeline. The \textit{ipyrad} pipeline consists of 7 steps, each performing one aspect of RADseq data processing \autocite{Eaton2020}. Our pseudo-individual of \textit{C. richardii} performed nearly identically to all of the other samples until Step 6. Here, similar sequences are clustered across samples, and then these shared sequences are mapped to a reference (our reference was a \textit{de novo} assembly, because the quality and coverage of the \textit{C. richardii} genome was too poor to serve as a reference). This requires that all individuals are related relatively close to one another, so that they share a large number of sequences. Very few sequences were retained for our pseudo-individual of \textit{C. richardii}, indicating that the sequences obtained are too different from the \textit{in vivo} samples. In order to retain enough SNPs to run STRUCTURE, we required that only 3 individuals needed to have data at a given locus for it to be included in the final data set. In our entirely \textit{in vivo} dataset, this minimum number of individuals was 30; we designated such a high number to reduce missing data. In our dataset with the pseudo-individual of \textit{C. richardii}, a vast majority of our retained loci had missing data.

The differences between the pseudo-individual derived from the \textit{Ceratopteris richardii} and our other samples is mostly likely due to the fact that the current assembly (v. 1.1) only represents about one third of the total genome, and consists of mostly short reads. Long reads that can span the many repetitive elements in the genome comprise only 0.03\% of the final assembly \autocite{Marchant2019}. Because of these factors, the current genome assembly contains a small amount of short sequences that are difficult to accurately compare to the fragments generated by \textit{in vivo} RADseq. We have included the results from our \textit{in silico} RADseq digest in Supplementary data 2, but do not present them in the main manuscript.

@@

To the best of our knowledge, this is the first application of next-generation sequencing techniques to the genus \textit{Ceratopteris}. Using population genomic, phylogenetic, and ploidy inferences, we recover similar species relationships found in previous studies, find support for \textit{C. gaudichaudii} as a genetically distinct cryptic species, identify a novel cryptic lineage from within \textit{C. thalictroides} \textit{sensu latu} in Central and South America, and show strong genetic variation across the broad range of \textit{C. coruta}.

Not included in our results are specimens of the model fern \textit{Ceratopteris richardii}, including the lab strain \textit{Hnn}. All individuals samples we had of this species yielded very poor sequencing results. We felt it was appropriate to remove these samples because we did not want to try and infer anything about this important species with poor quality data. Future work will be aimed at including wild collections of \textit{C. richardii} as well as the lab strain \textit{Hnn} to determine the relationship of this species to the rest of the genus.

@@

\textit{Ceratopteris cornuta} is found from western Africa to Northern Australia. The sterile fronds are consistently deltoid to lanceolate in shape, and pinnate to bipinnate in dissection \autocite{LloydTax1974}. Our population genomic analyses distinguished all specimens of \textit{C. cornuta} as a distinct population (at \textit{K} greater than 4, Fig. \ref{structure}). All samples of \textit{C. cornuta} grouped together at \textit{K} = 2 - 7, along with one sample of \textit{C. thalictroides} from Nepal (\textit{Fraser-Jenkins 1564}) and an unidentified sample from Brazil (\textit{L. Camargo de Abreu 21}).

Interestingly, our population genomic analysis and phylogenetic reconstruction show that individuals of \textit{C. cornuta} from the western portion of its range (Sierra Leon, Brazil) are genetically different from individuals in the eastern portion of its range (Tanzania, Oman, Nepal) (Fig. \ref{structure, phy}). At \textit{K} = 6, individuals from Sierra Leon (Cr04) and Brazil (Sp01) are separated slightly from the other three individuals of \textit{C. cornuta}, and contain some genetic influence that is not attributed to another sampled species. In addition, this small clade is placed as sister to the clade containing \textit{C. thalictroides sensu latu} and \textit{C. gaudichaudii} (Fig. \ref{phy}) It is important to note that there are no known collections of \textit{C. cornuta} from the Americas. However, \textit{Ceratopteris richardii} has been found in western Africa \cite{LloydTax1974}. We may have sampled African and South American individuals of \textit{C. richardii}. Similar patterns of Africa-South American gene flow have been seen in other broadly distributed pteridophytes. Recent gene flow between the two continents was observed in cosmopolitan \textit{Pteridium} \autocite{Wolf2019}, phylogenetic reconstruction has shown multiple migrations of grammitid ferns from South America to Africa \autocite{Sundue2014}, and several species (e.g. \textit{Asplenium monanthes}, \textit{Polypodium polypodioides} )have a disjunct distribution between the two continents \cite{kornas1993}. Therefore, it is possible that there could be some trans-Atlantic gene flow in \textit{Ceratopteris richardii}. Unfortunately, we do not have a known sample of \textit{C. richardii} to confidently compare to this group. When we did include samples of \textit{C. richardii} (from an herbarium specimen and the simulate digest of the \textit{C. richardii} genome) in a {\small{STRUCTURE}} analysis, they were placed roughly in the same population as \textit{C. cornuta} (Supplementary data 2). As we have mentioned before, these data were poor quality and low coverage. While it is a window into potential relationships, increased sampling is needed to draw any meaningful conclusions.

@@

Endemic to the New World tropics, \textit{Ceratopteris pteridoides} has perhaps the most unique morphology of the genus: sterile fronds are simple, palmately or pinnately lobed, with enlarged, air-filled stipes (petioles). They are most commonly found unrooted and floating on the surface of slow-moving water \autocite{LloydTax1974}. \textit{Ceratopteris pteridoides} is also very strongly differentiated by the {\small{STRUCTURE}} analysis, with samples identified as \textit{C. pteridoides} showing little to no introgression from other populations. Influence from \textit{C. pteridoides} does appear in \textit{C. cornuta}, \textit{C. gaudichaudii}, and some samples of \textit{C. thalictroides}. \textit{Ceratopteris pteridoides} is not known outside of the neotropics, and so it is difficult to say whether this is true introgression into other species, or if it is an artifact of slightly lower sequencing coverage in these individuals. It is known, however, that \textit{C. pteridoides} readily hybridizes with \textit{C. richardii} where their ranges overlap in northern South America \autocite{hickok1974}. Including the latter in future work may help to reveal the extent to which natural hybridization is occurring between diploid \textit{C. pteridoides} and \textit{C. richardii} in the New World.

@@

There was one named cryptic species of \textit{C. thalictroides} included in our study, \textit{Ceratopteris gaudichaudii}, which has two varieties: \textit{C. g. gaudichaudii} and \textit{C. g. vulgaris} \autocite{Masuyama2010}; only the latter was included in our analysis. \textit{Ceratopteris g. vulgaris} can be found throughout the Pacific in Japan, the Philippines, Hawaii, Guam, Taiwan, Southeast Asia, and Northern Australia; \textit{Ceratopteris g. gaudichaudii} is narrowly endemic to Guam \autocite{Masuyama2010}. Individuals identified as \textit{C. thalictroides} from Taiwan, Hawaii, Japan, China, and Australia were identified via our population genomic and phylogenetic analyses as being \textit{C. gaudichaudii}, although it is impossible to tell from our limited sampling which sub-species they align with. The individual from Australia was examined and keyed out morphologically to \textit{C. g. vulgaris}, using the key from Masuyama and Watano (2010). This same individual was growing a few meters away from an individual of Old World \textit{C. thalictroides}, indicating that these two morphologically similar species are growing sympatrically but with limited to no gene flow.

All of the samples from Hawaii included in this study show alignment to \textit{C. gaudichaudii} in our population structure analysis. \textcite{Wagner1950} hypothesized that \textit{Ceratopteris} was not native to Hawaii, and may have been introduced from Asia either as a food source or accidentally as a weed in taro patches. In addition, work by \textcite{Hickok1979} supported Hawaiian \textit{Ceratopteris} as a distinct species from \textit{C. thalictroides}. Japan was hypothesized to be the source of Hawaiian \textit{Ceratopteris} by \textcite{Lloyd1973}. This is plausible since \textit{Ceratopteris gaudichaudii} is relatively common in taro patches in Japan, and could have been brought to Hawaii via settlers or trade. Interestingly, one sample from Hawaii was shown to be a potential hybrid between \textit{C. gaudichaudii} and \textit{C. thalictroides}. The specimen (\textit{L. M. Crago 2005-058} US) is highly dissected and has may proliferous buds, which could indicate a hybrid or perhaps just an older plant. This sample was also placed as sister to Old World \textit{C. thalictroides} in our TETRAD and RAxML trees. Hybrid ancestry would explain grouping with \textit{C. gaudichaudii} in one analysis, and \textit{C. thalictroides} in another.

@@

The Old World \textit{C. thalictroides} clade includes mostly individuals from Asia and Australia, but also a few from South America. Increasing \textit{K} in {\small{STRUCTURE}} did not break up this large group, suggesting that there is recent gene flow across the Pacific via long distance dispersal or human-mediated transport. Several fern and angiosperm genera share such a disjunct distribution between eastern Asia and North to Central America, either by vicariance or long distance dispersal \autocite{Kato1983, Les2003}. One individual from Hawaii (\textit{L. M. Crago 2005-058} US) shows shared ancestry with \textit{C. gaudichaudii} and Old World \textit{C. thalictroides}, indicating that trans-Pacific gene flow might be possible in \textit{Ceratopteris}. Another individual from Japan shows a similar pattern, but with a much larger proportion of its genome comprised of the Old World \textit{C. thalictroides} population. There may be continued gene flow not only within Old World \textit{C. thalictroides}, but between \textit{C. thalictroides} and \textit{C. gaudichaudii}. The Hawaiian archipelago, as well as other Pacific islands, have been hypothesized to be a stepping-stone for trans-Pacific dispersal in angiosperm taxa \autocite{Harbaugh2009, Wright2001}, a theory that could also hold in ferns, which are known to have very high dispersal capabilities \autocite{Barrington1993, Tryon1970}.

>>>>

Cryptic species are present throughout the tree of life. They are especially prevalent in ferns, because of processes such hybridization, polyploidy, and reticulate evolution. In addition, the simple morphology of ferns limits phenotypic variation and makes it difficult to detect cryptic species. The model fern genus \textit{Ceratopteris} has long been suspected to harbor cryptic diversity, specifically in the highly polymorphic \textit{C. thalictroides}. Yet no studies have included samples from throughout its pan-tropical range or utilized genomic sequencing, making it difficult to assess the full extent of cryptic variation within this genus. Here, we present the first multilocus phylogeny of the genus using reduced representation genomic sequencing (RADseq) and examine population structure, phylogenetic relationships, and ploidy level variation. We recover similar species relationships found in previous studies, find support for the cryptic species \textit{C. gaudichaudii} as genetically distinct, and identify a novel genomic variation within two of the mostly broadly distributed species in the genus, \textit{C. thalictroides} and \textit{C. cornuta}.

@@

Despite the recent shift in molecular approaches to investigate cryptic species, much is still unknown about their evolutionary history and ecosystem functions. Cryptic species complexes often have very broad distributions \autocite{Der2009, Knowlton1993, Nygren2014} and can occur in sympatry, allopatry, or parapatry \autocite{Bickford2007}, making adequate sampling challenging. Studies have shown that cryptic species are prevalent across the Tree of Life (e.g. \cite{Bickford2007, Del_Carmen_Molina2011, Hebert2004, Nygren2014}), and ferns in particular are known to harbor many lineages with cryptic variation \autocite{Adjie2007, Paris1989, Yatabe2009}. However, deciphering species boundaries in ferns can be quite difficult due to their ease of dispersal via spores, making gene flow possible across vast distances \autocite{Barrington1993, Tryon1970}. Compared to seed plants, ferns have a high prevalence of polyploidy and reticulate evolution \autocite{Barrington1989, Paris1989, Sigel2016, Otto2000}. In addition, changes in ploidy can contribute to cryptic variation in ferns by altering niche space or offspring viability \autocite{Otto2007, Southgate2019, Masuyama2002}, yet have small effects on phenotype \autocite{Patel2019}. One example is the widespread fern genus \textit{Ceratopteris} Brong. (Pteridaceae), which is understood to have considerable cryptic variation, hybridization and polyploidy \autocite{Adjie2007, LloydTax1974, Masuyama2010}, making it an ideal system to investigate the origins and ecology of cryptic species.

@@

Our initial sampling contained 89 specimens of \textit{Ceratopteris} and one sample of \textit{Acrostichum aureum}, plus six replicates. After checking for quality and similarity of replicates, they were combined with the corresponding samples. Of the 90 unique starting samples of \textit{Ceratopteris}, 26 either extracted very poorly or yielded few loci (less than 1000 loci per individual). These were removed from downstream analysis. We also removed 13 samples from Australia, an area we were able to sample on a very fine spatial scale: one individual per Australian population was included in the final dataset so we did not over represent Australian specimens in our analyses. Unfortunately, all samples of \textit{Ceratopteris richardii} (including \textit{Hnn}) either failed in sequencing, or had very poor sequencing quality. After an initial examination of the results of these samples, we chose to remove them from our final analyses. When \textit{C. richardii} (simulated and/or from herbarium specimens) was included in our dataset, it increased the amount of missing data from 26\% to 80\%. \textcite{Hou2016} found that large amounts of missing data in RADseq studies can have a negative affect on phylogenetic reconstruction. We did not want to attempt to make any inferences about this important species, nor the lab strain, with so much missing data. Our final data set included 50 samples of \textit{Ceratopteris} and one sample of \textit{A. aureum}.

We did two separate runs of the \textit{ipyrad} pipeline: one with \textit{A. aureum} for phylogenetic analysis, and one without for ploidy estimation and population structure analysis. Because \textit{A. aureum} is sister to \textit{Ceratopteris} and separated by over 30 million years \autocite{PPGI}, the total number of SNPs retrieved for this individual was quite low.

@@

Due to the presence of numerous cryptic species and morphological variation across the genus, we hypothesize there may be cryptic cytotype variation and/or ploidy type (allo vs. auto) within species, especially the polymorphic \textit{C. thalictroides}. Since a majority our specimens were from herbarium accessions, we could not perform chromosome squashes. We utilized the R v. 3.5.2 \autocite{R_352} package gbs2ploidy to infer ploidy variation across samples. This package was designed to detect ploidy levels in variable cytoptype populations (specifically quaking aspen, \textit{Populus tremuloides}) using low-coverage (2X) genotyping-by-sequencing (GBS) or RADseq data; it was also tested on simulated data, and can detect diploids, triploids, or tetraploids \autocite{Gompert2017}. This package estimates ploidy from allele ratios, which are calculated from genome-average heterozygosity and bi-allelic SNPs isolated from a variant calling format (VCF) file, which is included in the output from \textit{ipyrad}.

@@

@@

To determine the best \textit{K} value across {\small{STRUCTURE}} we combined several metrics. We used the best \textit{K} method by \textcite{Evanno2005}, which determined a \textit{K} value of 4. The best \textit{K} estimate described in the {\small{STRUCTURE}} manual \autocite{Pritchard2000} determined a \textit{K} value of 5. We also visually examined \textit{K} = 2 - 6 and determined that \textit{K} = 5 was the most biologically meaningful because it separated each named species, and also showed some intra-specific variability (Fig. \ref{structure}). Increasing \textit{K} past 5 did not add any meaningful population clusters.

@@

 The phylogeny resulting from our species quartet inference shows a similar pattern to the {\small{STRUCTURE}} output (Fig. \ref{phy}). Individuals of \textit{Ceratopteris pteridoides} form a monophyletic group, sister to the rest of the genus \textit{Ceratopteris cornuta} came out in two places in the phylogeny: sister to Old World \textit{C. thalictroides}, and in a grade outside the \textit{C. gaudichaudii}, Old World and New World \textit{C. thalictroides} clade. Our phylogeny grouped all individuals placed in the \textit{C. gaudichaudii} population by {\small{STRUCTURE}} in a monophyletic clade, sister to New World \textit{C. thalictroides}.

 The {\small{RAxML}} tree shows a very similar topology to the {\small{TETRAD}} tree. The only major difference was that instead of New World \textit{C. thalictroides} and \textit{C. gaudichaudii} forming a monophyletic group, the former comes out as sister to all other species except \textit{C. pteridoides}. Since these results are nearly identical, we present a graphical comparison of the two trees in Supplementary data 2.

@@

Three of four samples identified as \textit{Ceratopteris cornuta} showed relatively equal allelic proportions of 1:1 and 3:1, indicating a possible tetraploid; the fourth sample was ambiguous. Some samples of the known diploid \textit{C. pteridoides} showed a similar pattern, with one putative diploid and three tetraploids. However, both species are known as a diploid so this may be a mis-assignment of ploidy by our analysis, or detecting of some interesting genomic structure. \textit{Ceratopteris gaudichaudii} had a high 1:1 allele ratio, suggesting that it is an allotetraploid. \textit{Ceratopteris thalictroides} is also known to be a tetraploid \autocite{LloydTax1974, Masuyama2010}, and individuals in our analysis were estimated to have 1:1, 2:1, and 3:1 allele ratios. Despite this variation, there was a distinct pattern between Old World and New World individuals: Old World \textit{C. thalictroides} was most commonly had a 1:1 allele ratio (autopolyploid), whereas New World \textit{C. thalictroides} had a 3:1 allele ratio (allopolyploid).

@@

We attempted to create a pseudo-individual to include in our analysis using the \textit{Ceratopteris richardii} genome assembly. We were able to obtain 9,047,554 simulated raw sequences utilizing an \textit{in silico} RADseq digest. These reads, in combination with the rest of our \textit{in vivo} data, were processed with the \textit{ipyrad} pipeline. When constructing the final assembly, \textit{ipyrad} requires that all individuals are related relatively similar to one another, so that they share a large number of sequences. Very few sequences were retained for our pseudo-individual of \textit{C. richardii}, indicating that the sequences obtained are too different from the \textit{in vivo} samples. In order to retain enough SNPs to run STRUCTURE, we required that only 3 individuals needed to have data at a given locus for it to be included in the final data set. In our entirely \textit{in vivo} dataset, this minimum number of individuals was 30; we designated such a high number to reduce missing data. In our dataset with the pseudo-individual of \textit{C. richardii}, 80\% of our retained loci had missing data.

The differences between the pseudo-individual derived from the \textit{Ceratopteris richardii} and our other samples is mostly likely due to the fact that the current assembly (v. 1.1) only represents about one third of the total genome, and consists of mostly short reads. Long reads that can span the many repetitive elements in the genome comprise only 0.03\% of the final assembly \autocite{Marchant2019}. Because of these factors, the current genome assembly contains a small amount of short sequences that are difficult to accurately compare to the fragments generated by \textit{in vivo} RADseq.

To explore the utility of our simulated data, we ran a population structure analysis \autocite{Pritchard2000} for this new dataset including the simulated \textit{C. richardii} and one additional specimen of the same species (see Supplementary data 2). This dataset had 80\% missing data, compared to 26\% without \textit{C. richardii}. Both of these individuals had mixed ancestry from several different species in the genus, which is not what we were expecting: \textit{Ceratopteris richardii} has been assumed to be a diploid species, and although it hybridizes with other species in the genus, it was not thought to be of hybrid origin itself. The individual derived from the \textit{C. richardii} genome aligned closely with \textit{C. cornuta}; this is interesting because the two species ranges’ overlap in western Africa, and they have very similar morphology. We can certainly speculate about potential hybrid origins or mis-identifications of specimens, but without higher quality data, we are hesitant to make any major claims. The population structure of all other species in the genus remained virtually identical to our analysis including only the highest-quality individuals (as seen in Fig. \ref{structure}). We have included the data simulation pipeline and the population structure analysis including \textit{C. richardii} in the Supplementary data 2.

@@

To the best of our knowledge, this is the first application of next-generation sequencing techniques to the genus \textit{Ceratopteris}. Using population genomic, phylogenetic, and ploidy inferences, we recover similar species relationships found in previous studies, find support for \textit{C. gaudichaudii} as a genetically distinct cryptic species, identify a novel cryptic lineage from within \textit{C. thalictroides} \textit{sensu latu} in Central and South America, and show strong genetic variation across the broad range of \textit{C. cornuta} (Fig. \ref{map}).

Not included in our results are specimens of the model fern \textit{Ceratopteris richardii}, including the lab strain \textit{Hnn}. All individuals samples we had of this species yielded very poor sequencing results. When included in our analyses, they significantly increased the amount of missing data. We felt it was appropriate to remove these samples because we did not want to try and infer anything about this important species with poor quality data. Future work will be aimed at including wild collections of \textit{C. richardii} as well as the lab strain \textit{Hnn} to determine the relationship of this species to the rest of the genus.

@@

\textit{Ceratopteris cornuta} is found from western Africa to Northern Australia. The sterile fronds are consistently deltoid to lanceolate in shape, and pinnate to bipinnate in dissection \autocite{LloydTax1974}. Our population genomic analyses distinguished all specimens of \textit{C. cornuta} as a distinct population (at \textit{K} greater than 4, Fig. \ref{structure}). All samples of \textit{C. cornuta} grouped together at \textit{K} = 2 - 6, along with one sample of \textit{C. thalictroides} from Nepal (\textit{Fraser-Jenkins 1564}) and an unidentified sample from Brazil (\textit{L. Camargo de Abreu 21}).

Interestingly, our population genomic analysis and phylogenetic reconstruction show that individuals of \textit{C. cornuta} from the western portion of its range (Sierra Leon, Brazil) are genetically different from individuals in the eastern portion of its range (Tanzania, Oman, Nepal) (Fig. \ref{structure, phy}). At \textit{K} = 6, individuals from Sierra Leon (Cr04) and Brazil (Sp01) are separated slightly from the other three individuals of \textit{C. cornuta}, and contain some genetic influence that is not attributed to another sampled species. In addition, this small clade is placed as sister to the clade containing \textit{C. thalictroides sensu latu} and \textit{C. gaudichaudii} (Fig. \ref{phy}). It is important to note that there are no known collections of \textit{C. cornuta} from the Americas. However, \textit{Ceratopteris richardii} has been found in western Africa \autocite{LloydTax1974}. We may have sampled African and South American individuals of \textit{C. richardii}. The morphology of the two species is somewhat similar \autocite{LloydTax1974}, and so these two specimens may have been mis-identified by their collectors. Similar patterns of Africa-South American gene flow have been seen in other broadly distributed pteridophytes. Recent gene flow between the two continents was observed in cosmopolitan \textit{Pteridium} \autocite{Wolf2019}, phylogenetic reconstruction has shown multiple migrations of grammitid ferns from South America to Africa \autocite{Sundue2014}, and several species (e.g. \textit{Asplenium monanthes}, \textit{Polypodium polypodioides}) have a disjunct distribution between the two continents \autocite{kornas1993}. Therefore, it is possible that there could be some trans-Atlantic gene flow in \textit{Ceratopteris richardii}. Unfortunately, we do not have a known sample of \textit{C. richardii} to confidently compare to this group. When we did include samples of \textit{C. richardii} (from an herbarium specimen and the simulated digest of the \textit{C. richardii} genome) in a {\small{STRUCTURE}} analysis, they were placed roughly in the same population as \textit{C. cornuta} (Supplementary data 2). As we have mentioned before, these data were poor quality and low coverage. While it is a window into potential relationships, increased sampling is needed to draw any meaningful conclusions.

@@

Endemic to the New World tropics, \textit{Ceratopteris pteridoides} has perhaps the most unique morphology of the genus: sterile fronds are simple, palmately or pinnately lobed, with enlarged, air-filled stipes (petioles). They are most commonly found unrooted and floating on the surface of slow-moving water \autocite{LloydTax1974}. \textit{Ceratopteris pteridoides} is also very strongly differentiated by the {\small{STRUCTURE}} analysis, with samples identified as \textit{C. pteridoides} showing little to no introgression from other populations. Influence from \textit{C. pteridoides} does appear in \textit{C. cornuta}, \textit{C. gaudichaudii}, and some samples of \textit{C. thalictroides}. It is difficult to say whether this is true introgression into other species, or if it is an artifact of slightly lower sequencing coverage in these individuals. It is known, however, that \textit{C. pteridoides} readily hybridizes with \textit{C. richardii} where their ranges overlap in northern South America \autocite{hickok1974}. Including the latter in future work may help to reveal the extent to which natural hybridization is occurring between diploid \textit{C. pteridoides} and \textit{C. richardii} in the New World.

@@

There was one named cryptic species of \textit{C. thalictroides} included in our study, \textit{Ceratopteris gaudichaudii}, which has two varieties: \textit{C. g. gaudichaudii} and \textit{C. g. vulgaris} \autocite{Masuyama2010}; only the latter was included in our analysis. \textit{Ceratopteris g. vulgaris} can be found throughout the Pacific in Japan, the Philippines, Hawaii, Guam, Taiwan, Southeast Asia, and Northern Australia; \textit{Ceratopteris g. gaudichaudii} is narrowly endemic to Guam \autocite{Masuyama2010}. Individuals identified as \textit{C. thalictroides} from Taiwan, Hawaii, Japan, China, and Australia were identified via our population genomic and phylogenetic analyses as being \textit{C. gaudichaudii}, although it is impossible to tell from our limited sampling which subspecies they align with. The individual from Australia was examined and keyed out morphologically to \textit{C. g. vulgaris}, using the key from Masuyama and Watano (2010). This same individual was growing a few meters away from an individual of Old World \textit{C. thalictroides}, indicating that these two morphologically similar species are growing sympatrically but with limited to no gene flow.

All of the samples from Hawaii included in this study show alignment to \textit{C. gaudichaudii} in our population structure analysis. \textcite{Wagner1950} hypothesized that \textit{Ceratopteris} was not native to Hawaii, and may have been introduced from Asia either as a food source or accidentally as a weed in taro patches. In addition, work by \textcite{Hickok1979} supported Hawaiian \textit{Ceratopteris} as a distinct species from \textit{C. thalictroides}. Japan was hypothesized to be the source of Hawaiian \textit{Ceratopteris} by \textcite{Lloyd1973}. This is plausible since \textit{Ceratopteris gaudichaudii} is relatively common in taro patches in Japan, and could have been brought to Hawaii via settlers or trade. Interestingly, one sample from Hawaii was shown to be a potential hybrid between \textit{C. gaudichaudii} and \textit{C. thalictroides}. The specimen (\textit{L. M. Crago 2005-058} US) is highly dissected and has many proliferous buds, which could indicate a hybrid or perhaps just an older plant. This sample was also placed as sister to Old World \textit{C. thalictroides} in our TETRAD and RAxML trees. Hybrid ancestry would explain grouping with \textit{C. gaudichaudii} in one analysis, and \textit{C. thalictroides} in another.

@@

The Old World \textit{C. thalictroides} clade includes mostly individuals from Asia and Australia, but also a few from South America (Fig. \ref{map}). Increasing \textit{K} in {\small{STRUCTURE}} did not break up this large group, suggesting that there is recent gene flow across the Pacific via long distance dispersal or human-mediated transport. Several fern and angiosperm genera share such a disjunct distribution between eastern Asia and North to Central America, either by vicariance or long distance dispersal \autocite{Kato1983, Les2003}. One individual from Hawaii (\textit{L. M. Crago 2005-058} US) shows shared ancestry with \textit{C. gaudichaudii} and Old World \textit{C. thalictroides}, indicating that trans-Pacific gene flow might be possible in \textit{Ceratopteris}. Another individual from Japan shows a similar pattern, but with a much larger proportion of its genome comprised of the Old World \textit{C. thalictroides} population. There may be continued gene flow not only within Old World \textit{C. thalictroides}, but between \textit{C. thalictroides} and \textit{C. gaudichaudii}. The Hawaiian archipelago, as well as other Pacific islands, have been hypothesized to be a stepping-stone for trans-Pacific dispersal in angiosperm taxa \autocite{Harbaugh2009, Wright2001}, a theory that could also hold in ferns, which are known to have very high dispersal capabilities \autocite{Barrington1993, Tryon1970}.

New World \textit{C. thalictroides} is strongly differentiated from Old World \textit{C. thalictroides} in our population structure analysis, and shows virtually no gene flow between any other species at \textit{K} 4 - 6. This strong distinction between Old and New World is present in the phylogeny as well: New World \textit{C. thalictroides} is placed in a clade as sister to \textit{C. gaudichaudii}, and this clade is in turn sister to Old World \textit{C. thalictroides} (Fig. \ref{phy}). \textit{Ceratopteris gaudichaudii} has been hypothesized to be an allopolyploid hybrid \autocite{Adjie2007}; perhaps the derivation of its sister species, New World \textit{C. thalictroides}, could have been from a hybridization event with one of the parents of \textit{C. gaudichaudii}, or genome doubling to form an autotetraploid, followed by long distance dispersal. Our ploidy inference suggests that either form of polyploidy is possible, but slightly more individuals of New World \textit{C. thalictroides} appeared to be of allopolyploid origin, potentially with subsequent genomic rearrangement (Fig. \ref{g2p}).
